---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

---
title: "Documentation For hdjmtTest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hdjmtTest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `hdjmtTest` package performs high-dimensional mediator screening and testing in the presence of high correlation among mediators using Ridge-HOLP and Approximate Orthogonalization methods. In addition to accounting for correlation, the method is designed to detect mediators that exhibit joint effects. This vignette demonstrates how to use the primary functions with a simulated dataset, `ExampleData`.

# Structure of ExampleData

The example data provided with the package contains:

- information on `n = 200` samples and `p = 2000` potential active mediators.
- **Variables**:
  - `M`: A 200x2000 matrix of mediators generated using the compound symmetry covariance structure to introduce high correlation among the mediators, with $\rho= 0.8$.
  - `x`: A vector of length 200 representing exposures.
  - `y`: A vector of length 200 representing outcomes.
  - `alp_vec`: A parameter vector of length 200 that relates the exposure variable to the mediators. The first 8 values are `-0.5127127, 0.6597036, 0.6756640, 0.5235137, 0.9305369, -0.9827865, -0.8941141, -0.9230220` with the rest being zeros.
  - `beta_vec`: A parameter vector of length 200 that relates the mediators to the outcome variable. The first 12 values are `-0.8033093, 0.9360975, 0.8185305, -0.7951502, -0.8783739, 0.8940459, -0.6911509, -0.8524771, -0.6812097, -0.8285034, -0.5986530, -0.9639383` with the rest being zeros.
- **Active Mediators**: given the non-zero values in `alp_vec` and `beta_vec`, only the first 8 mediators are truly active.

# How to install package from github

```{r}
#library(devtools)

#devtools::install_github('samuelOsarfo/hdjmtTest')
```

If package "qvalue" is not found, please first install "qvalue" package through Bioconductor: https://www.bioconductor.org/packages/release/bioc/html/qvalue.html

```{r}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("qvalue")
```

# Summary of Functions

## `medsc_holp`

- **Description**: Performs mediator screening using the Ridge-HOLP method.
- **Usage**: `medsc_holp(y, x, M, d = NULL, r = 1)`
- **Arguments**:
  - `y`: Outcome vector.
  - `x`: Exposure vector.
  - `M`: Matrix of mediators.
  - `COV.S`:  a `data.frame` or  `matrix` of covariates (optional).
  - `d`: Desired number of mediators to select (optional). Default value is $d= 0.5\cdot n/\log(n)$.
  - `r`: Ridge-HOLP penalty parameter (default is 1).

## `app_orth`

- **Description**: Fits the Approximate Orthogonalization model to test the significance of mediators.
- **Usage**: `app_orth(y, x, chosen_M, COV.S=NULL, k = 1)`
- **Arguments**:
  - `y`: Outcome vector.
  - `x`: Exposure vector.
  - `chosen_M`: Matrix of mediators selected during screening.
  - `COV.S`:  a `data.frame` or  `matrix` of covariates (optional).
  - `k`: Scalar used to compute projection directions (default is 1).

## `get_active_med.hima`, `get_active_med.mod`, `get_active_med.hdmt`

- **Description**: These are wrapper functions that combine screening and testing to identify active mediators using different joint-significance test variants.
- **Usage**: `get_active_med.<variant>(y, x, M, COV.S=NULL, pval.adjust='HDMT', d=NULL, r=1, k=1)`
- **Variants**:
  - `get_active_med.hima`: Uses the `null_estimation` function from the original HIMA repository.
  - `get_active_med.mod`: Uses a modified version of the `null_estimation` function.
  - `get_active_med.hdmt`: Uses the implementation in the HDMT package.

- **Arguments**:
  - `y`: Outcome vector.
  - `x`: Exposure vector.
  - `M`: Matrix of mediators.
  - `COV.S`:  A `data.frame` or `matrix` of covariates (optional).
  - `pval.adjust`: Specifies which method to use for controlling FWER in the joint significance testing. Either `HDMT` (default) or `Bonferroni`.
  - `d`: Desired number of mediators to select (optional). Default is $d=  n/\log(n)$.
  - `r`: Ridge penalty for HOLP (default = 1).
  - `k`: Scalar for computing projection directions for AO (default = 1).

```{r}
# Load the ExampleData
library(hdjmtTest)
data(ExampleData)

# Extract the components
y <- ExampleData$y
x <- ExampleData$x
M <- ExampleData$M
```

# Mediator Screening using Ridge-HOLP

```{r}
# Perform Ridge-HOLP screening
chosen_ind <- medsc_holp(y, x, M)

print("Indexes of selected mediators:")
print(chosen_ind)
```

# Fitting Approximate Orthogonalization (AO)

```{r}
# Apply AO on the chosen mediators
chosen_med <- M[, medsc_holp(y, x, M)]
ao_result <- app_orth(y, x, chosen_med)

print("Test statistics for selected mediators:")
print(ao_result$ts)

print("P-values for selected mediators:")
print(ao_result$pval)
```

# Identifying Active Mediators

```{r}
# Using HDMT (HIMA null_estimation)
active_mediators.hima <- get_active_med.hima(y, x, M)
print(active_mediators.hima)

# Using HDMT (modified null_estimation)
active_mediators.mod <- get_active_med.mod(y, x, M)
print(active_mediators.mod)

# Using HDMT package implementation
active_mediators.hdmt <- get_active_med.hdmt(y, x, M)
print(active_mediators.hdmt)

# Using Bonferroni correction
active_mediators_Bonferroni <- get_active_med.hima(y, x, M, pval.adjust='bonferroni')
print(active_mediators_Bonferroni)
```

# Competing packages

```{r}
## Install the HIMA package
# install.packages('HIMA')

suppressMessages(library(HIMA))

HIMA::hima_dblasso(x, M, y)
```

Out of the 8 active mediators, `M[, 1], M[,3], M[,4], M[,5], M[,6], M[,7], M[,8]` were identified as active by HIMA (specifically HIMA2).

# Conclusion

The `hdjmtTest` package simplifies high-dimensional mediation analysis, providing tools for selecting active mediators under high correlation scenarios. Refer to the function documentation for more details.

# Reference

- Wang, X., & Leng, C. (2016). High dimensional ordinary least squares projection for screening variables. *Journal of the Royal Statistical Society Series B: Statistical Methodology*, 78(3), 589–611.
- Battey, H. S., & Reid, N. (2023). On inference in high-dimensional regression. *Journal of the Royal Statistical Society Series B: Statistical Methodology*, 85(1), 149–175.
