---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

---
title: "Documentation For hdjmtTest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hdjmtTest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Introduction

The `hdjmtTest` package performs high-dimensional mediator screening and testing in the presence of  high correlation among mediators using Ridge-HOLP and Approximate Orthogonalization methods. This vignette demonstrates how to use the primary functions with a simulated dataset, `ExampleData`.


# Structure of ExampleData

The example data provided with the package  contains:

- information for `200` observations and `2000` potential active mediators.
- **Variables**:
  - `M`: A 200x2000 matrix of mediators generated using the compound symmetry covariance structure to introduce high correlation among the mediators, with $\rho= 0.8$.
  - `x`: A vector of length 200 representing exposures.
  - `y`: A vector of length 200 representing outcomes.
  - `alp_vec`: A parameter vector of length 200 that relates the exposure variable to the mediators. The first 8 values are `-0.5127127, 0.6597036, 0.6756640, 0.5235137, 0.9305369, -0.9827865, -0.8941141, -0.9230220` with the rest being zeros.
  - `beta_vec`: A parameter vector of length 200 that relates the mediators to the outcome variable. The first 12 values are  `-0.8033093, 0.9360975, 0.8185305, -0.7951502, -0.8783739, 0.8940459, -0.6911509, -0.8524771, -0.6812097, -0.8285034,           -0.5986530, -0.9639383` with the rest being zeros.
- **Active Mediators**: given the non-zero values in `alp_vec` and `beta_vec`, only the first 8 mediators are truly active




# How to install package from github

```{r}
#library(devtools)

#devtools::install_github('samuelOsarfo/hdjmtTest')

```

If package "qvalue" is not found, please first install "qvalue" package through Bioconductor: https://www.bioconductor.org/packages/release/bioc/html/qvalue.html

```{r}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("qvalue")

```


# Summary of Functions

## `medsc_holp`

- **Description**: Performs mediator screening using the Ridge-HOLP method.
- **Usage**: `medsc_holp(y, x, M, d = NULL, r = 1)`
- **Arguments**:
  - `y`: Outcome vector.
  - `x`: Exposure vector.
  - `M`: Matrix of mediators.
  - `COV.S`:  a `data.frame` or  `matrix` of covariates (optional).
  - `d`: Desired number of mediators to select (optional). Default value is $d= 0.5\cdot n/\log(n)$.
  - `r`: Ridge-HOLP penalty parameter (default is 1).

## `app_orth`

- **Description**: Fits the Approximate Orthogonalization model to test the significance of mediators.
- **Usage**: `app_orth(y, x, chosen_M, COV.S=NULL, k = 1)`
- **Arguments**:
  - `y`: Outcome vector.
  - `x`: Exposure vector.
  - `chosen_M`: Matrix of mediators selected during screening.
  - `COV.S`:  a `data.frame` or  `matrix` of covariates (optional).
  - `k`: Scalar used to compute projection directions (default is 1).

## `get_active_med`

- **Description**: A wrapper function that combines screening and testing to identify active mediators.
- **Usage**: `get_active_med(y, x, M, COV.S=NULL,pval.adjust='HDMT',d=NULL, r=1, k=1)`
- **Arguments**:
  - `y`: Outcome vector.
  - `x`: Exposure vector.
  - `M`: Matrix of mediators.
  - `COV.S`:  A `data.frame` or `matrix` of covariates (optional).
  - `pval.adjust`: Specifies which method to use for controlling FWER in the joint significance testing. Either `HDMT` (default)               or `Bonferroni`.
  - `d`:  Desired number of mediators to select (optional). Default value is $d= 0.5\cdot n/\log(n)$.
  - `r`:  Penalty parameter for the Ridge-HOLP. Default value is `1`
  - `k` : Scalar used to compute projection directions for AO. Default value is `1`.



 

```{r}
# Load the ExampleData
library(hdjmtTest)
data(ExampleData)

# Extract the components
y <- ExampleData$y
x <- ExampleData$x
M <- ExampleData$M


```

# Mediator Screening using Ridge-HOLP

The `medsc_holp` function performs Ridge-HOLP screening to select potential active mediators based on their association with the outcome and exposure.

```{r}
# Perform Ridge-HOLP screening
chosen_ind <- medsc_holp(y, x, M)


print("Indexes of selected mediators:")
print(chosen_ind)

```


# Fitting Approximate Orthogonalization (AO)
With the selected mediators, we can use `app_orth` to fit the Approximate Orthogonalization model. This model computes test statistics and p-values for the selected mediators.

```{r}
# Apply AO on the chosen mediators
chosen_med <- M[, medsc_holp(y, x, M)]
ao_result <- app_orth(y, x, chosen_med)


# Print the test statistics and p-values
print("Test statistics for selected mediators:")
print(ao_result$ts)

print("P-values for selected mediators:")
print(ao_result$pval)

```


# Identifying Active Mediators
The `get_active_med` function integrates screening and testing, identifying active mediators and their corresponding indexes. The steps involved are as follows:

### 1. Mediator Screening with Ridge-HOLP
The function starts by using the Ridge-HOLP method to reduce the dimensionality of the mediator matrix (`M`). This step ensures that only potential mediators with strong associations with both the exposure (`x`) and the outcome (`y`) are considered for further testing.

### 2. Fitting the AO Model
After initial screening, the function fits the Approximate Orthogonalization (AO) model to the selected mediators. This fitting produces test statistics and p-values that help determine the significance of each mediator.

### 3. Joint-Significance Test
- A joint-significance test is performed to identify active mediators using the p-values obtained from both the AO model and the mediator model.

- To control for the family-wise error rate due to multiple testing, Bonferroni/HDMT correction is applied to the p-values.

- **Output**:  Indexes of the mediators that are deemed significant after the joint-significance test.

```{r}
# Identify active mediators

#Using HDMT in joint significance testing (Default)
active_mediators_HDMT <- get_active_med(y, x, M) 

#Indexes of active mediators identified using HDMT:"
print(active_mediators_HDMT)



#Using Bonferroni in joint significance testing
active_mediators_Bonferroni <- get_active_med(y, x, M, pval.adjust='bonferroni') 

#Indexes of active mediators identified using Bonferroni:"
print(active_mediators_Bonferroni)
```



# Competing packages
We can perform this test using `HIMA` package as well.

```{r}
## Install the HIMA package
#install.packages('HIMA')

suppressMessages(library(HIMA))

```

The `dblassoHIMA` function in the HIMA package combines the sure independence screening method and de-biased lasso to identify the active mediators. Consider the rcode below for identifying the active mediators in the simulated data.


```{r}

HIMA::dblassoHIMA(x,M,y)
```
Out of the 8 active mediators, `M[, 1], M[,3], M[,4], M[,5], M[,6], M[,7], M[,8]` were identified as active my HIMA (specifically HIMA2).




# Conclusion

The `hdjmtTest` package simplifies high-dimensional mediation analysis, providing tools for selecting active mediators under high correlation scenarios. Refer to the function documentations for more details.


# Reference

- Wang, X., & Leng, C. (2016). High dimensional ordinary least squares projection for screening variables. Journal of the Royal Statistical Society Series B: Statistical Methodology, 78(3), 589-611.

- Battey, H. S., & Reid, N. (2023). On inference in high-dimensional regression. Journal of the Royal Statistical Society Series B: Statistical Methodology, 85(1), 149-175.
